On Optimality of Finite Element Integration

### 2015-10-31 TOMS_Paper/assembly.tex
Again, trying to improve section 2. non si vede la fine

In this section, we study optimal \textit{sharing elimination}; that is, a transformation that minimizes a loop nest operation count by applying factorization, code motion and (if necessary) common sub-expression elimination. In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies. However, we show that by restricting ourselves to the class of loops and expressions in Definition~\ref{def:fem-loopnest}, the search space becomes very small.

Finite element integration expressions can be reduced to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the problem, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. Note that this respects Definition~\ref{def:fem-loopnest}. We make the following considerations:

\begin{enumerate}
\item In the special case in which test and trial functions are in the same function space, there exists $w1$ and $w2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$. That is, common sub-expressions arise over different iteration spaces ($L_j$ and $L_k$). By hoisting $\alpha$ or $\beta$ at the level of the closest common loop ($L_i$), the operation count is reduced.
\item Let us now consider the general case in which test and trial functions belong to different function spaces. By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed (a simplistic example is showed in Figure~\ref{code:multi_loopnest}). Similarly to point 1), such a factorization aims to minimize the operation count ``locally'' at the level of the outer product loop nest. This transformation is independent of the chosen function spaces, but it may not be optimal in case point 1) is applicable (a common sub-expression elimination opportunity is lost). Two observations follow:
\begin{enumerate}
\item In case both 1) and 2) are applicable, there is no simple way of predicting what transformation provides the largest decrease in operation count; 
\item However, the difference between the two tends to be negligible, since the outer product loop nests eventually perform the same number of operations.
\end{enumerate}
\item If we neglect the pathological cases in which the loops degenerate (i.e., when the trip count becomes 1, the case of P0 discontinuous elements), points 1) and 2) always result in reduced operation count.
\item Factorizing any of the symbols in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) also impacts the operation count. We distinguish two cases:
\begin{enumerate}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form can be reduced to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$. 
\item There are no common symbols among the terms. Factorizing any $f_i \in \sigma_{ei}$ would then

\item The starting point of the factorization is the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$. Since the symbols in $\sigma_{ei}$ are generally not common factors

In this case, the production of non-factorizable, non-hoistable sub-expressions tends to drastically inhibit code motion. In the best case (e.g., there is a single scalar coefficient pre-multiplying the form) this simply reduces to cases 1) or 2). In other words, this is a ``non-natural factorization'', which tries to group potentially non-common factors and does not take advantage of loop dependencies.
\item Point 1) was applied -- for example, we have $a_{ij}(b_{ik}(\sigma'_{ei} + \sigma''_{ei}))$, with $\sigma_{ei} = \sigma'_{ei} + \sigma''_{ei}$ -- and we factorize symbols within $\sigma_{ei}$. Splitting $L_e$-dependent from $L_i$-dependent sub-expressions (through suitable re-factorization) and applying code motion naturally tends to operation count minimization. Similarly to point 3), however, with very low trip counts this strategy may be sub-optimal. For example, consider the following sub-expression, $(a v_i + b w_i) (c v_i + d w_i)$, which is excerpted from a multiplication of the inverse Jacobian matrix for the change of coordinates with the derivative of a vector-valued coefficient. It is true that factorizing $v_i$ and $w_i$ exposes $L_i$-independent sub-expressions, but this may increase the operation count when the size of $L_i$ is 1.
\end{enumerate}
\end{enumerate}

Algorithm~\ref{algo:opt} turns these observations into a sequence of steps leading to optimal sharing elimination.

\begin{Algo}[Sharing elimination]
\label{algo:opt}
\normalfont
Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$. The algorithm starts with reducing the input expression to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$. The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are then ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree. We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms. The choice concerning what factorize first, which impacts the resulting operation count, depends on two factors: the number of unique terms in each loop, $n$, and the trip counts, $tc$. It is fairly simple to see (and easily demonstrable by contradiction) that by starting from the loop such that $n tc$ is smallest leads to the maximal operation count reduction. Assume then we start with $L_j$. Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression. The other sub-expression, independent of $L_j$ ($L_k$), is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting. Finally, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted, checking that this actually results in reduced operation count (see point 4b) above).
\end{Algo}

>>>>

In this section, we study optimal \textit{sharing elimination}; that is, a transformation that minimizes a loop nest operation count by applying factorization, code motion and (if necessary) common sub-expression elimination. In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies. However, we show that by restricting ourselves to the class of loops and expressions in Definition~\ref{def:fem-loopnest}, the search space is actually very small.

Finite element integration expressions can be reduced to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ (the number of ``terms'') depends on the problem, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. Note that this normal form respects Definition~\ref{def:fem-loopnest}. We use the normal form for the following considerations:

\begin{enumerate}
\item In the special case in which test and trial functions are in the same function space, there exists $w1$ and $w2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$. That is, common sub-expressions arise over different iteration spaces ($L_j$ and $L_k$). By hoisting $\alpha$ or $\beta$ at the level of the closest common loop ($L_i$), the operation count is reduced.
\item Let us now consider the general case in which test and trial functions belong to different function spaces. By factorizing, in sequence, any $a_{ij} \in \alpha_{eij}$ and then any $b_{ik} \in \beta_{eik}$ (or vice-versa), sub-expressions independent of one of the outer product loops are exposed (a simplistic example is showed in Figure~\ref{code:multi_loopnest}). Similarly to point 1), such a factorization aims to minimize the operation count ``locally'' at the level of the outer product loop nest. This transformation is independent of the chosen function spaces, but it may not be optimal in case point 1) is applicable (a common sub-expression elimination opportunity is lost). Two observations follow:
\begin{enumerate}
\item In case both 1) and 2) are applicable, there is no simple way of predicting what transformation provides the largest decrease in operation count; 
\item However, the difference between the two tends to be negligible, since the operation count within the outer product loop nest is identical.
\end{enumerate}
\item Factorizing any of the symbols in $\sigma_{ei}$ (e.g., a coefficient, or a coefficient derivative along a certain spatial dimension) also impacts the operation count. We distinguish two cases:
\begin{enumerate}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$. 
\item There is no common symbol among terms. Then, factorizing any $f_i \in \sigma_{ei}$ \textit{across two terms} $w1$ \textit{and} $w2$ always increases the operation count with respect to points 1) or 2). This is because of two facts. Firstly, the number of operations at every iteration is larger. Secondly, sub-expressions that are not hoistable due to the inherent dependencies on loop indices are exposed. On the other hand, factorizing any $f_i \in \sigma_{ei}$ \textit{within a generic term} $w$ promotes hoisting of geometry-dependent sub-expressions at the level of $L_e$. 
\end{enumerate}
\item Inherent in the previous points is the fundamental fact that there never are two ore more conflicting factorizations among which to choose for maximizing code motion. The fact that expressions result from sequences of inner products and loop nests adhere to a specific structure allows us to define a rigid transformation process leading, in general, to operation count minimization. 
\item There only are a few corner circumstances in which point 4) decades. Briefly:
\begin{enumerate}
\item As already explained, the dichotomy between points 1) and 2).
\item The case in which loops degenerate (i.e., their trip counts is 1), which makes code motion irrelevant and certain factorizations sub-optimal (e.g., the factorization $(a f_i + b g_i) (c f_i + d f_i)$ is optimal if and only if the size of $L_i$ is 1).
\end{enumerate}
\end{enumerate}

These considerations can easily be turned into a constructive proof showing that any loop nest produced by Algorithm~\ref{algo:opt} is optimal if and only if: (i) we restrict, as specified at the beginning of the section, to factorization, code motion and common sub-expression elimination (what we mean by this is also clarified in the next section); (ii) we neglect the pathological cases in point 5) above.

\begin{Algo}[Sharing elimination]
\label{algo:opt}
\normalfont
Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$. The algorithm starts with reducing the input expression to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$. The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are then ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree. We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms. The choice concerning what factorize first, which impacts the resulting operation count, depends on two factors: the number of unique terms in each loop, $n$, and the trip counts, $tc$. It is fairly simple to see (and easily demonstrable by contradiction) that by starting from the loop such that $n \cdot tc$ is smallest leads to the maximal operation count reduction. Assume then we start with $L_j$. Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression. The other sub-expression, independent of $L_j$ ($L_k$), is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting. Finally, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted, checking that this actually results in reduced operation count (see point 4b) above).
\end{Algo}
